A multi-scale story of the diffusion of a new technology: the web

### 2025-01-03 paper/submission1/anonymised.tex
ssh problem

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A multi-scale story of the diffusion of a new technology: the web},
  pdfauthor={Emmanouil Tranos},
  pdfkeywords={keyword1, keyword2},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{A multi-scale story of the diffusion of a new technology: the
web}
\author[1]{Emmanouil Tranos%
\corref{cor1}%
}
 \ead{e.tranos@bristol.ac.uk}

\affiliation[1]{organization={University of
Bristol},postcode={UK},postcodesep={}}

@@

\begin{abstract}
This paper maps the participation in the digital economy and its
evolution in the UK over space and time. Most of the existing economic
geography literature which dealt with the spatiality of the internet
employed supply-side measures, such as infrastructural capacity, in
order to understand the geography of the digital economy and its
potential spatial economic effects. Useful as these approaches might
have been, they cannot capture the micro-processes and the
characteristics of the individual online behaviour. Using large volumes
of archived and geolocated web content, this paper models the diffusion
of web technologies over space and time in the UK. The data and
geolocation strategy allow to capture these processes at a very granular
spatial scale. The modelling approach, which is based on simple spatial
analytical methods and on the estimation of diffusion curves at various
scales, enables to depict the role of geography and other cognitive
factors which drove the diffusion of web technologies. Although the
focus is on a recent historical period -- 1996-2012 -- the results of
the analysis depict diffusion mechanisms which can be very useful in
understanding the evolutionary patterns of the adoption of other newer
technologies.
\end{abstract}

\begin{keyword}
    keyword1 \sep 
    keyword2
\end{keyword}
\end{frontmatter}

@@

Economic geographers were always interested in how new technologies and
innovations diffuse across space and time with the work of
\citet{hagerstrand1968innovation}, which demonstrated the importance of
communications as a mechanism for the diffusion of innovations, being
the most impactful. It was followed by various empirical studies -- see
for instance \citet{ormrod1990} and \citet{iso2005}. However, the torch
of exploring and modelling such diffusion processes had been passed to
other disciplines such as economics, business studies and sociology well
before the `cultural turn' of economic geography
\citep{perkins2005international}. A potential explanation of the recent
lack of geographical studies exploring the diffusion of new and, more
specifically for this paper, digital technologies across both space and
time can be attributed to the scarcity of relevant and granular enough
data, a problem also highlighted by \citet{iso2005} and
\citet{kemeny2011international}. As \citet{zook2022mapping} pointed,
digital activities are hardly ever captured in official data.

This paper offers such a contribution: a geographical study illustrating
how a new technology that is the Web diffused over space and time in the
UK at a high level of spatial granularity during the 1996-2012 period.
It does so by employing a novel source of big data and machine learning
algorithms to model the \emph{active} engagement with Web during that
period and tests how different spatial and temporal mechanisms shaped
its diffusion. This paper exemplifies how the combination of data
sources which escape the traditional social science domain and adequate
research methods can offer new lenses to geographical research regarding
the understanding of technological diffusion.

The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffused over space and time and use this knowledge to make predictions
about the diffusion of related \emph{future} technologies. As per
\citet{leibowicz2016representing}, historical studies agree that
technologies diffuse differently in terms of times, rates, and
geographies and can be driven by related policies \citep{victor1993}.
\citet{meade2021modelling} highlight that a variety of actors have a
direct interest in gaining such knowledge including network equipment
suppliers, network operators, regulatory and local authorities. These
processes and their effects vary a lot across scales: although the
diffusion of a new technology might not be optimal at a local level, it
might be beneficial from a global perspective as it could lead to faster
diffusion to less advantaged places \citep{leibowicz2016representing}.
Despite the spatial heterogeneity of such diffusion mechanisms and the
policy relevance, there is a scarcity of empirical studies analysing the
diffusion of new digital technologies at a detailed geographical level.

This gap becomes even more prominent if we consider the importance of
technology and its adoption in economic geography and macroeconomic
growth theories. Simply put, differences in technological adoption and
sophistication may lead to uneven development as new technologies can
increase productivity and allow firms to introduce new products
\citep{solow1957technical, aghion1990model, kemeny2011international}.
Linked to the subject of this paper, the Web -- and the active
engagement with this technology that is creating instead of just
browsing a website -- can be considered as the underpinning technology
for the transformative process known as the \emph{digital service
economy} \citep{capello2024nexus}. After all, diffusion together with
invention and innovation are considered the pillars of technological
change \citep{das2022diffusion}. Findings from previous empirical work
illustrated that early engagement with digital technologies such as the
Web led to longstanding positive productivity gains for these regions
that hosted early adopters \citep{tranosuk}.

Another line of motivation has to do with the intangible nature of the
Web and the debate about the footloose nature of digital technologies,
which contrary to several earlier general purpose technologies are light
in weight \citep{coyle1999weightless}. Following the seminal work of
\citet{gaspar1998information}, which demonstrated that technological
improvements in telecommunications lead to increased demand for
face-to-face communications and, consequently, enhance the importance of
cities, empirical studies from economic geography and urban economics
explored whether digital technologies substitute or complement
agglomeration externalities and spatial structure
\citep{kolko_death_2000, ioannides2008effect, tranos2021ubiquitous}.
With the exception of \citet{sinai2004geography}, who focused, among
other things, on web content and demonstrated that more such content is
available in larger markets, this strand of literature has been mostly
concerned with the spatiality of the internet's hard and tangible
infrastructure and shied away from exploring the spatial dimensions of
one of the most distinct software layers of the internet infrastructure,
the Web.

@@

This paper focuses on the diffusion of the Web as new technology during
the 1996-2012 period. This was an exciting period for digital
technologies as it corresponds with the commercialisation of the
internet and, consequently, its almost universal adoption. The reader is
reminded that it was only in 1994 when Netscape Navigator was
introduced, a year before Microsoft's Internet Explorer.\footnote{\url{https://www.theguardian.com/global/2015/mar/22/web-browser-came-back-haunt-microsoft}}
Also, only 9 per cent of UK's households had access to the internet in
1998 \citep{ons2018}, the web included mostly static webpages, there
were no social media and web browsing involved exclusively a desktop PCs
as there were no smartphones \citep{tranosuk}. Hence, it is fair to say
that the study period captures the very early stages of the diffusion of
a new technology that is the Web as well until its maturity. The former
is a key point in the lifecycle of a new technology. Firstly, during
this period new technologies are expensive, crude and imperfect
\citep{rosenberg1994exploring, wilson201281}. A simple comparison
between Web 1.0 and Web 3.0 applications clearly illustrates this
\citep{tranos2020social}: while a static website facilitates one-way
information dissemination, a platform like \texttt{github} enables
cooperation between users and the creation of new information, meaning,
and even knowledge \citep{faraj2016special, barassi2012does}. During
this period the performance of a new technology is the main attraction
and not the cost to access and use it \citep{wilson2011lessons}. There
is a broader theoretical discussion in the literature about the early
adoption motives. As summarised by \citet{perkins2005international}, on
one hand, epidemic models highlight the role of interpersonal contacts
as a way for new technologies to diffuse. On the other hand, economic
models underline the importance of heterogeneity. Different firms have
different structures and business plans, which define the potential
economic returns of the adoption of a new technology and, therefore, the
choice to adopt a new technology becomes an individual option. From a
broader and evolutionary perspective, initial conditions are essential
for the creation and evolution of path-dependent technological
development trajectories \citep{neffke2011regions, simmie2014new}. This
argument is even more relevant when the focus is on digital technologies
because of the commonly found lag between investment and economic
returns as reflected in the Solow paradox
\citep{acemoglu2014return, brynjolfsson2018artificial}.

The data used here depict the active engagement with the Web as they
contain geolocated and time-stamped counts of commercial websites from
the UK. Instead of adopting metrics reflecting the passive engagement
with the internet such as internet adoption and speeds
\citep[e.g.][]{blank2018local, destefano2022fuzzy}, this paper observes
yearly counts of geolocated websites, which reflect the creation and
maintenance of websites. It needs to be highlighted here that although
there are various types of underpinning web technologies -- namely
markup, stylesheet, client- and server-side coding languages as well as
different web browsers and other related programming language -- the
focus here is on the diffusion of the Web in its totality. For a more
detailed and cross-country technological diffusion study, see
\citet{PAPAGIANNIDIS2015308}.

To analyse these data, a diverse set of methods are employed. Firstly,
cumulative adoption curves are estimated for the UK as a whole and for
the Local Authority Districts (LAD), illustrating, for the first time,
such S-shaped diffusion patterns for local areas. This system level
analysis also reveals insights about the volatility of the adoption of
the Web. Then, exploratory spatial data analysis offers some first
indications of the importance of the different spatial diffusion
mechanisms in shaping the diffusion of the Web in the UK. Importantly,
the analysis adopts two distinct geographical scales: the LAD and the
much more detailed scale of the smallest census areas in the UK, the
Output Areas. Lastly, a novel methodological framework based on
well-established Machine Learning (ML) algorithms is developed to test
the role of such mechanisms at these two scales. The results reveal for
the first time the importance and relevance of different spatial
diffusion mechanisms for the diffusion of a tangible, digital technology
at very granular geographical scales.

The structure of the paper goes as follows. Section~\ref{sec-litreview}
reviews the key concepts in the diffusion literature as well as relevant
empirical studies. Then, Section~\ref{sec-datamethods} describes the
data and methods used in this paper. The results of the different stages
of the analysis are presented in Section~\ref{sec-results}. The
discussion of the results and the conlcusions of the paper are offered
in Section~\ref{sec-conclusions}.

\textbf{SAY SOMETHING ABOUT DIST PAPER BY ARP, NOT SPECIFIC EXPLANATORY
VARIABLES}

@@

Geographical diffusion is a synthesis of different processes. On the one
hand, a purely spatial or, in other words, contagious processes can be
identified. Adjacency and, more broadly, distance are the key drivers of
diffusion. This perspective draws similarities with epidemics:
innovation just like pathogens spreads because of contagion and,
consequently, proximity and exposure \citep{hivner2003facilitating}. On
the other hand, there is a hierarchical processes. Instead of a
horizontal distance-based diffusion mechanism, the top-down hierarchy of
urban systems shapes technological diffusion. In reality, the synthesis
of these two processes better reflects how new technologies diffuse over
space and time \citep{morrill2020spatial}.

These ideas were firstly introduced by Torsten Hägerstrand and his
thesis entitled `Innovation Diffusion as a Spatial Process'
\citep{hagerstrand1968innovation}. Hägerstrand was the first one to
identify diffusion as a geographical process. The starting point was the
idea that diffusion is based on passing information through social
networks, which themselves tend to be defined by geography. Hence, he
identified a `neighbourhood' mechanism of how information, and
consequently, innovation diffuse. He used agricultural innovations to
test and model his ideas using Monte Carlo simulations. Hägerstrand also
incorporated the role of hierarchy and how some phenomena maybe firstly
adopted in larger cities and then diffuse to second tier ones. This is a
sequential instead of a simultaneous process, which resembles the
`lead-lag' spatial acceleration effect in market research
\citep{bento2018time, PERES201091}. Hägerstrand is more widely known
though for highlighting the role time plays in the diffusion of
innovations: an early-pioneering period, a middle fast accelerating
period and a final saturation one \citep{morrill2020spatial}.

@@

\citet{grubler1990rise} effectively summarises the three key spatial
diffusion mechanisms: (i) the cumulative level of adoption follows an
S-shaped pattern just like purely temporal models; (ii) diffusion is
shaped by a hierarchy mechanisms in a form of a centrifugal force: from
core to periphery; and (iii) diffusion is also shaped by distance and
proximity. These are the three mechanisms that the empirical analysis in
Section~\ref{sec-results} investigates.

The remaining of this section reviews empirical studies which analysed
the diffusion on new technologies over space and time. Although the
spatial dimension is present in most of the following studies, the level
of spatial detail is always more coarse than the one adopted in this
paper. \citet{beardsell1999spatial} studied the evolution of the
computer industry in 317 US metro areas during the 1977-1992 period
using employment data. Their analysis indicated that the relative size
distribution holds for urban computer employment and also urban
heterogeneity is essential in explaining this distribution. In a recent
study, \citet{bednarz2020pulled} focused on wind turbines and modeled
their spatial diffusion across 402 German regions during 1970-2015.
Their key finding is that local demand than local supply was the main
driving factor. \citet{haller2011determinants} employed firm panel data
for Ireland during 2003-2005 and illustrated that when firms are located
in a region or industry with high share of firms having a website, they
are more likely to have a website too.

At a global scale \citet{perkins2005international} explored whether the
diffusion rate of new technologies is driven by a latecomer advantage
and the engagement with the global economy via foreign direct
investments and trade. Their results illustrate that indeed latecomers
and developing countries experience diffusion of new technologies more
rapidly than early adopters and developed countries. At the same scale,
\citet{perkins2011internet} explored whether the adoption of previous
communication technologies that is mail, telegrams and telephones were
shaped by similar socioeconomic factors as the internet. Their results
indicated common patterns regarding the drivers behind the adoption of
different communication technologies.

Turning to studies that share more technological and scalar similarities
with this paper, \citet{ding2010modeling} modeled the spatial diffusion
of mobile telecommunications across regions in China. Their analysis
indicated that socioeconomic characteristics are important determinants
of the timing, speed and the level of mobile diffusion within China.
Using data from a Hungarian online social network,
\citet{lengyel2020role} analysed its adoption and the churn at a very
granular spatial level. Their results are in agreement with early
theoretical and empirical contributions reviewed here: assortativity,
urban scaling and distance are the key drivers of spatial diffusion. At
a global scale \citet{PAPAGIANNIDIS2015308} modeled the diffusion of
different web technologies and practices across countries.
Interestingly, they used similar, but less extensive data as the ones
used here. Their analysis illustrated how the diffusion of different web
technologies and practices follow an S-shaped pattern as well as the
different diffusion rates of the different technologies and practices.

All in all, this paper tests whether the three distinct spatial
diffusion mechanisms shaped the diffusion of the Web in the UK at very
granular spatial scales from the Web's early stages onwards.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Journal Name}

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A multi-scale story of the diffusion of a new technology: the web},
  pdfkeywords={diffusion, technologies, Web, digital},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{A multi-scale story of the diffusion of a new technology: the
web}

@@

\begin{abstract}
Despite the broad consensus that differences in technological adoption
and sophistication can lead to uneven trajectories of local development
as new technologies can increase productivity and allow firms to
introduce new products and services, economic geographers have shied
away from exploring the diffusion of digital technologies across space
and time. This paper fills this gap by exposing the roles of different
spatial mechanisms in the diffusion of an intangible, digital technology
at very granular geographical scales. It does so by employing a novel
source of data and machine learning algorithms to model the active
engagement with Web during 1996-2012.
\end{abstract}

\begin{keyword}
    diffusion \sep technologies \sep Web \sep 
    digital
\end{keyword}
\end{frontmatter}

@@

Economic geographers were always interested in how new technologies and
innovations diffuse across space and time with the work of
\citet{hagerstrand1968innovation}, which demonstrated the importance of
communications as a mechanism for the diffusion of innovations, being
the most impactful. It was followed by various empirical studies -- see
for instance \citet{ormrod1990} and \citet{iso2005}. However, the torch
of exploring and modelling such diffusion processes had been passed to
other disciplines such as economics, business studies and sociology well
before the `cultural turn' of economic geography
\citep{perkins2005international}. A potential explanation of the lack of
geographical studies exploring the diffusion of new and, more
specifically for this paper, digital technologies across both space and
time can be attributed to the scarcity of relevant and granular enough
data, a problem also highlighted by \citet{iso2005} and
\citet{kemeny2011international}. As \citet{zook2022mapping} pointed,
digital activities are hardly ever captured in official data.

This paper offers such a contribution: an economic geography study
revealing the importance of spatial mechanisms for the diffusion of an
intangible, digital technology that is the Web in the UK at a high level
of spatial granularity during the 1996-2012 period. It does so by
employing a novel source of big data and machine learning algorithms to
model the \emph{active} engagement with Web during that period and tests
how different spatial and temporal mechanisms shaped its diffusion. This
paper also exemplifies how the combination of data sources and
state-of-the-art research methods which escape the traditional social
science domain can offer new lenses to geographical research regarding
the understanding of technological diffusion.

The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffuse over space and time and use this knowledge to make predictions
about the diffusion of related future technologies. As per
\citet{leibowicz2016representing}, historical studies agree that
technologies diffuse differently in terms of times, rates, and
geographies and can be driven by related policies \citep{victor1993}.
\citet{meade2021modelling} highlight that a variety of actors have a
direct interest in gaining such knowledge including network equipment
suppliers, network operators, regulatory and local authorities. These
processes and their effects vary a lot across scales: although the
diffusion of a new technology might not be optimal at a local level, it
might be beneficial from a global perspective as it could lead to faster
diffusion to less advantaged places \citep{leibowicz2016representing}.
Despite the spatial heterogeneity of such diffusion mechanisms and the
policy relevance, there is a scarcity of studies analysing the diffusion
of new digital technologies at a detailed geographical level because of
the lack of detailed enough data and adequate research methods.

This gap becomes even more prominent if we consider the importance of
technology and its adoption in economic geography and macroeconomic
growth theories. Simply put, differences in technological adoption and
sophistication may lead to uneven development as new technologies can
increase productivity and allow firms to introduce new products
\citep{solow1957technical, aghion1990model, kemeny2011international}.
Linked to the subject of this paper, the Web -- and the active
engagement with this technology that is creating and maintaining instead
of just browsing a website -- can be considered as the underpinning
technology for the transformative process identified by
\citet{capello2024nexus} as the \emph{digital service economy}.
Importantly, the focus here is not the Web in general, but instead
commercial websites. After all, diffusion together with invention and
innovation are considered the pillars of technological change
\citep{das2022diffusion}. Findings from previous studies illustrated
that early engagement with digital technologies such as the Web led to
longstanding positive productivity gains for these regions that hosted
early adopters \citep{tranosuk}. Hence, obtaining a robust understanding
of the spatial diffusion mechanisms is crucial in mapping future
development trajectories.

Another line of motivation has to do with the intangible nature of the
Web and the debate about the footloose nature of digital technologies,
which contrary to several earlier general purpose technologies are
`light in weight' \citep{coyle1999weightless}. Following the seminal
work of \citet{gaspar1998information}, which demonstrated that
technological improvements in telecommunications lead to increased
demand for face-to-face communications and, consequently, enhance the
importance of cities, empirical studies from economic geography and
urban economics explored whether digital technologies substitute or
complement agglomeration externalities and spatial structure
\citep{kolko_death_2000, ioannides2008effect, tranos2021ubiquitous}.
With the exception of \citet{sinai2004geography}, who focused, among
other things, on web content and demonstrated that more such content is
available in larger markets, this strand of literature has been mostly
concerned with the spatiality of the Internet's hard and tangible
infrastructure and shied away from exploring the spatial dimensions of
one of the most distinct software layers of the Internet infrastructure,
the Web.

@@

This paper focuses on the diffusion of the Web as new technology during
the 1996-2012 period. This was an exciting period for digital
technologies as it corresponds with the commercialisation of the
Internet and, consequently, its almost universal adoption. The reader is
reminded that it was only in 1994 when Netscape Navigator was
introduced, a year before Microsoft's Internet Explorer.\footnote{\url{https://www.theguardian.com/global/2015/mar/22/web-browser-came-back-haunt-microsoft}}
Also, only 9 per cent of UK's households had access to the Internet in
1998 \citep{ons2018}, the web included mostly static webpages, there
were no social media and web browsing involved exclusively a desktop PCs
as there were no smartphones \citep{tranosuk}. Hence, it is fair to say
that the study period captures the very early stages of the diffusion of
a new technology that is the Web until its maturity. The former is a key
point in the lifecycle of a new technology. Firstly, during this period
new technologies are expensive, crude and imperfect
\citep{rosenberg1994exploring, wilson201281}. A simple comparison
between Web 1.0 and Web 3.0 applications clearly illustrates this
\citep{tranos2020social}: while a static website facilitates one-way
information dissemination, a platform like \texttt{github} enables
cooperation between users and the creation of new information, meaning,
and even knowledge \citep{faraj2016special, barassi2012does}. During
this period the performance of a new technology is the main attraction
and not the cost to access and use it \citep{wilson2011lessons}. There
is a broader theoretical discussion in the literature about the early
adoption motives. As summarised by \citet{perkins2005international}, on
one hand, epidemic models highlight the role of interpersonal contacts
as a way for new technologies to diffuse. On the other hand, economic
models underline the importance of heterogeneity. Different firms have
different structures and business plans, which define the potential
economic returns of the adoption of a new technology and, therefore, the
choice to adopt a new technology becomes an individual option. From a
broader and evolutionary perspective, initial conditions are essential
for the creation and evolution of path-dependent technological
development trajectories \citep{neffke2011regions, simmie2014new}. This
argument is even more relevant when the focus is on digital technologies
because of the commonly found lag between investment and economic
returns as reflected in the Solow paradox
\citep{acemoglu2014return, brynjolfsson2018artificial}.

The data used here depict the active engagement with the Web as they
contain geolocated and time-stamped counts of commercial websites from
the UK. Instead of adopting metrics reflecting the passive engagement
with the Internet such as Internet adoption and speeds
\citep[e.g.][]{blank2018local, destefano2022fuzzy}, this paper observes
yearly counts of geolocated websites, which reflect the creation and
maintenance of websites. It needs to be highlighted here that although
there are various types of underpinning web technologies -- namely
markup, stylesheet, client- and server-side coding languages as well as
different web browsers and other related programming language -- the
focus here is on the diffusion of the Web in its totality. For a more
detailed and cross-country technological diffusion study, see
\citet{PAPAGIANNIDIS2015308}.

To analyse these data, a diverse set of methods are employed. Firstly,
cumulative adoption curves are estimated for the UK as a whole and for
the Local Authority Districts (LAD), illustrating, for the first time,
such S-shaped diffusion patterns for local areas. This system level
analysis also reveals insights about the volatility of the adoption of
the Web. Then, exploratory spatial data analysis offers some first
indications of the importance of the different spatial diffusion
mechanisms in shaping the diffusion of the Web in the UK. Importantly,
the analysis adopts two distinct geographical scales: the LAD and the
most detailed scale of UK census areas, the Output Areas (OA). Lastly, a
novel methodological framework based on Machine Learning (ML) algorithms
is developed to test the role of such mechanisms at these two scales.
The results reveal, for the first time, the importance of different
spatial diffusion mechanisms in the diffusion of an intangible, digital
technology at very granular geographical scales.

The structure of the paper goes as follows. Section~\ref{sec-litreview}
reviews the key concepts in the diffusion literature as well as relevant
empirical studies. Then, Section~\ref{sec-datamethods} describes the
data and methods used in this paper. The results of the different stages
of the analysis are presented in Section~\ref{sec-results}. The
discussion of the results and the conclusions of the paper are offered
in Section~\ref{sec-conclusions}.

@@

Geographical diffusion is a synthesis of different processes. On the one
hand, a purely spatial or, in other words, contagious processes can be
identified. Adjacency and, more broadly, distance are the key drivers of
diffusion. This perspective draws similarities with epidemics:
innovations just like pathogens spread because of contagion and,
consequently, proximity and exposure \citep{hivner2003facilitating}. On
the other hand, there is a hierarchical processes. According to this,
instead of a horizontal distance-based diffusion mechanism, the top-down
hierarchy of urban systems shapes technological diffusion. In reality,
the synthesis of these two processes better reflects how new
technologies diffuse over space \citep{morrill2020spatial}.

These ideas were firstly introduced by Torsten Hägerstrand and his
thesis entitled `Innovation Diffusion as a Spatial Process'
\citep{hagerstrand1968innovation}. Hägerstrand was the first to identify
diffusion as a geographical process. The starting point was the idea
that diffusion is based on passing information through social networks,
which themselves tend to be defined by geography. Hence, he identified a
`neighbourhood' mechanism of how information, and consequently,
innovation diffuse. He used agricultural innovations to test and model
his ideas using Monte Carlo simulations. Hägerstrand also incorporated
the role of hierarchy and how some phenomena maybe firstly adopted in
larger cities and then diffuse to second tier ones. This is a sequential
instead of a simultaneous process, which resembles the `lead-lag'
spatial acceleration effect in market research
\citep{bento2018time, PERES201091}. Hägerstrand is more widely known
though for highlighting the role time plays in the diffusion of
innovations: an early-pioneering period, a middle fast accelerating
period and a final saturation one \citep{morrill2020spatial}.

@@

\citet{grubler1990rise} effectively summarises the three key spatial
diffusion mechanisms: (i) the cumulative level of adoption follows an
S-shaped pattern just like purely temporal models; (ii) diffusion is
shaped by a hierarchical mechanism in a form of a centrifugal force:
from core to periphery; and (iii) diffusion is also shaped by distance
and proximity. These are the three mechanisms that the empirical
analysis in Section~\ref{sec-results} investigates.

The remaining of this section reviews empirical studies which analysed
the diffusion on new technologies over space and time. Although the
spatial dimension is present in most of the following studies, the level
of spatial detail is always more coarse than the one adopted in this
paper. \citet{beardsell1999spatial} studied the evolution of the
computer industry in 317 US metro areas during the 1977-1992 period
using employment data. Their analysis indicated that the relative size
distribution holds for urban computer employment and also urban
heterogeneity is essential in explaining this distribution. In a recent
study, \citet{bednarz2020pulled} focused on wind turbines and modelled
their spatial diffusion across 402 German regions during 1970-2015.
Their key finding is that local demand than local supply was the main
driving factor. \citet{haller2011determinants} employed firm panel data
for Ireland during 2003-2005 and illustrated that when firms are located
in a region or industry with high share of firms having a website, they
are more likely to have a website too.

At a global scale \citet{perkins2005international} explored whether the
diffusion rate of new technologies is driven by a latecomer advantage
and the engagement with the global economy via foreign direct
investments and trade. Their results illustrate that indeed latecomers
and developing countries experience diffusion of new technologies more
rapidly than early adopters and developed countries. At the same scale,
\citet{perkins2011internet} explored whether the adoption of previous
communication technologies that is mail, telegrams and telephones were
shaped by similar socioeconomic factors as the Internet. Their results
indicated common patterns regarding the drivers behind the adoption of
different communication technologies.

Turning to studies that share more technological and scalar similarities
with this paper, \citet{ding2010modeling} modelled the spatial diffusion
of mobile telecommunications across regions in China. Their analysis
indicated that socioeconomic characteristics are important determinants
of the timing, speed and the level of mobile diffusion within China.
Using data from a Hungarian online social network,
\citet{lengyel2020role} analysed its adoption and the churn at a very
granular spatial level. Their results were in agreement with the early
theoretical and empirical contributions reviewed here: assortativity,
urban scaling and distance are the key drivers of spatial diffusion.
\citet{bakher2013diffusion} compared different the growth curves as a
means to model diffusion of online shopping in Australia during the
1998-2009 period. At a global scale \citet{PAPAGIANNIDIS2015308}
modelled the diffusion of different web technologies and practices
across countries. Using similar but less extensive data as the one used
here, their analysis illustrated how the diffusion of different web
technologies and practices follow an S-shaped pattern as well as the
different diffusion rates of the different technologies and practices.

All in all, this paper tests whether the three distinct spatial
diffusion mechanisms shaped the diffusion of the Web in the UK at very
granular spatial scales from the Web's early stages onwards. The next
section discusses the empirical framework.
